1-intro

### 2022-04-19 1-intro.Rmd
some fix to intro

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and multiple variables measured at each spatial and temporal value. In data collection, spatio-temporal data is less often collected as a whole piece but fragmented spatial and temporal data for analysts to ensemble into spatio-temporal data. With the [integrated] data, we hope to fix the time to explore the spatial distribution, or fix the spaital location to explore temporal trends. Ideally, we would be able to carry out each of these tasks easily, and even accommodate complex explorations encompassing both simultaneously. These are all underpinned by a [flexible] data structure that can represent spatio-temporal data.

@@

What remains to be difficult when wrangling spatio-temporal data is that the raw data is not naturally structured in any one of the representation mentioned above and the first difficulty is to cast the data, potentially from multiple sources into one of the data representation. Some examples of these includes map data recorded the shape of a collection of area of interest is provided by the government website; geo-scientific data recorded with longitude and latitude coordinate of locations in the area; and temporal variables indexed by both location and time for the temporal aspect. For spatio-temporal data analysis, merging all these sources can be challenging and joining these tables can go wrong because of the slightest unmatched of the linking variable from different sources.

Another issue is how the concept of tidy data should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable; and 3) one type of data per table. The long form data is preferred over wide data given the downstream data cleaning and visualisation software i.e. \pkg{dplyr} and \pkg{ggplot2}. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over long time period, i.e. sub-daily time series data.

>>>>

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and multiple variables measured at each spatial and temporal value. The analysis of spatio-temporal data can fix the time to explore the spatial distribution, fix the spatial location to explore temporal trends, or dynamically explore the space and time simultaneously. For these tasks to be carried out easily, there is a need for a spatio-temporal data representation that underpins various operations.

@@

What remains to be difficult when wrangling spatio-temporal data is that the raw data is not naturally structured in any one of the representation above and the first task is to cast the data into a class. Also, in data collection, spatio-temporal data is less often collected as a whole piece but fragmented spatial and temporal data for analysts to ensemble. For example, map data records the shape of a collection of area of interest is provided by official government website; geo-scientific data records the longitude and latitude coordinate of locations in the area along with other meta data relates to the location; and temporal variables indexed by both location and time. For spatio-temporal data analysis, merging all these sources can be challenging and joining these tables can go wrong because of the slightest unmatched of the linking variable from different sources.

Another issue is how the concept of tidy data [@tidydata] should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable; and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e. \pkg{dplyr} and \pkg{ggplot2} for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over long time period. This poses the question on how to arrange spatial and temporal variables that makes analysing spatio-temporal data easier.
